DisjunctiveProgramming.jl: Generalized Disjunctive Programming Models and Algorithms for JuMP

### 2023-10-14 paper/paper.tex
fix #63
begin fixes to paper

\subsection{Linear GDP reformulation example}
The simplest example of a linear GDP system is given in \eqref{eq:ex} - \eqref{eq:y}, where $Y_i$ is a Boolean indicator variable that enforces the constraints in the disjunct ($Ax \le b$ or $Cx \le d$) when $true$.
\vskip 6pt
\begin{equation}
    \label{eq:ex}
    \begin{bmatrix}
    Y_1 \\ Ax \leq b
    \end{bmatrix}
    \lor
    \begin{bmatrix}
        Y_2 \\ Cx \leq d
    \end{bmatrix}
\end{equation}
\begin{equation}
    \label{eq:x}
    0 \leq x \leq U
\end{equation}
\begin{equation}
    \label{eq:simple_xor}
    Y_1 \ \underline{\vee} \ Y_2
\end{equation}
\begin{equation}
    \label{eq:y}
    Y_1, Y_2 \in \{true, false\}
\end{equation}
\vskip 6pt

@@

 \subsubsection{Big-M Reformulation}
 The Big-M reformulation for this problem is given by \eqref{eq:x}, \eqref{eq:ex_bigm1} - \eqref{eq:ex_bigm4}, where $M$ is a sufficiently large scalar that makes the particular constraint redundant when its indicator variable is not selected (i.e., $y_i = 0$). Note that the Boolean variables, $Y_i$, are replaced by binary variables, $y_i$. When the integrality constraint in Eq. \eqref{eq:ex_bigm4} is relaxed to $0 \leq x_1, x_2 \leq 1$, the resulting feasible region can be visualized by projecting the relaxed model onto the $x_1, x_2$ plane. This results in the region encapsulated by the dashed line in Figure \ref{fig:bigm}. It should be noted that the relaxed feasible region is not as tight as possible around the original feasible solution space. The choice of the large $M$ value determines the tightness of this relaxation, and the minimal value of $M$ for the optimal relaxation can be found through interval arithmetic when the model is linear. For nonlinear models, the tightest $M$ can be obtained by solving the maximization problem $\{\max h_{ik}(x): x \in X\}$. An alternate method for tight Big-M relaxations is given in \cite{TRESPALACIOS201598}.

\begin{equation}
    \label{eq:ex_bigm1}
    Ax \leq b + M \cdot (1 - y_1)
\end{equation}
\begin{equation}
    \label{eq:ex_bigm2}
    Cx \leq d + M \cdot (1 - y_2)
\end{equation}
\begin{equation}
    \label{eq:ex_bigm3}
    y_1 + y_2 = 1
\end{equation}
\begin{equation}
    \label{eq:ex_bigm4}
    y_1, y_2 \in \{0,1\}
\end{equation}

@@

\subsubsection{Hull Reformulation}
The Hull reformulation is given by \eqref{eq:x}, \eqref{eq:ex_bigm3} - \eqref{eq:ex_hull3}, which requires lifting the model to a higher-dimensional space. When projected to the original space, the continuous relaxation of the model is tighter than its Big-M equivalent \cite{grossmann_trespalacios_2013}. The reformulation relaxation can be visualized by the region encapsulated by the dashed line in Figure \ref{fig:chr}. Note that this reformulation provides a tighter relaxation than the Big-M reformulation in Figure \ref{fig:bigm}. Also note that describing the geometry of this relaxation is more complex than the Big-M relaxation, which is made possible by the increased number of constraints and variables in the model.

\begin{equation}
    \label{eq:ex_hull1}
    Ax_1 \leq by_1
\end{equation}
\begin{equation}
    \label{eq:ex_hull0}
    Cx_2 \leq dy_2
\end{equation}
\begin{equation}
    \label{eq:ex_hull2}
    x = x_1 + x_2
\end{equation}
\begin{equation}
    \label{eq:ex_hull3}
    0 \leq x_i \leq U y_i \quad \forall i \in \{1,2\}
\end{equation}

@@

Once the logic propositions are converted to CNF, each clause can be converted into an algebraic constraint with the following equivalence (Note: any negated Boolean variables, $\neg Y_i$, are replaced with $1-y_i$ in the reformulation),

\begin{align*}
    \bigvee_{i \in I} Y_i & \ \ \text{becomes} \ \ \sum_{i\in I} y_i \geq 1 \\
\end{align*}

@@

\begin{lstlisting}[language = Julia]
using DisjunctiveProgramming, JuMP, HiGHS

# create model
m = JuMP.Model(HiGHS.Optimizer)
# add variables to model
@variable(m, 0 <= F[i = 1:7] <= 10)
@variable(m, 0 <= CS <= CSmax)
@variable(m, CRmin <= CR <= CRmax)

# add constraints to model
@constraints(m,
    begin
        F[1] == F[2] + F[3]
        F[7] == F[5] + F[6]
    end
)
\end{lstlisting}
    \item Define the inner (nested) disjunction for the separation technologies in the superstructure using the \verb|@disjunction| macro.   
\begin{lstlisting}[language = Julia]
@disjunction(m,
    begin
        F[5] == β[:S1]*F[4]
        CS == γ[:S1]
    end,
    begin
        F[5] == β[:S2]*F[4]
        CS == γ[:S2]
    end,
    reformulation = :big_m, # reformulation type
    name = :YS # symbol for indicator variable
)
\end{lstlisting}
    \item Define constraints in the outer disjunctions.
\begin{lstlisting}[language = Julia]
# define constraints in left disjunct
R1_con = @constraints(m,
    begin
        F[6] == β[:R1]*F[2]
        [i = 3:5], F[i] == 0
        CR == γ[:R1]
        CS == 0
    end
)

# define constraints in right disjunct
R2_con = @constraints(m,
    begin
        F[6] == β[:R2]*F[3]
        CR == γ[:R2]      
    end
)
\end{lstlisting}
    \item Build the main disjunction using the constraint blocks defined in (3) and the \verb|add_disjunction!| function. Note that the reformulated constraints for the nested disjunction are stored in the \verb|.ext| dictionary of the model under the name of the disjunction (\verb|:YS| in this case).
\begin{lstlisting}[language = Julia]
add_disjunction!(m,
    R1_con,
    (
        R2_con, #general constraints in R2 disj.
        m.ext[:YS] #reformulated inner disj.
    ),
    reformulation = :big_m, # reformulation type
    name = :YR # symbol for indicator variable
)
\end{lstlisting}
    \item Add the selection logical constraints using the \verb|choose!| function. The first constraint enforces that only one reactor is selected (i.e., $Y_{R_1} \ \underline{\vee} \ Y_{R_2}$). The second constraint enforces that the separation system be defined only if the second reactor ($R_2$) is selected. This constraint is equivalent to the proposition $Y_{R_2} \Leftrightarrow Y_{S_1} \ \underline{\vee} \ Y_{S_2}$.
\begin{lstlisting}[language = Julia]
choose!(1, YR[1], YR[2]; mode = :exactly)
choose!(YR[2], YS[1], YS[2]; mode = :exactly)
\end{lstlisting}
    \item Add the objective function and optimize. 
\begin{lstlisting}[language = Julia]
@objective(m, Max, F[7] - CS - CR)
optimize!(m)
\end{lstlisting}
\end{enumerate}

>>>>

\subsection{Solution Technique: Reformulation to Mixed-Integer Program}
The simplest example of a linear GDP system is given in \eqref{eq:ex} - \eqref{eq:y}, where $Y_i$ is a Boolean indicator variable that enforces the constraints in the disjunct ($Ax \le b$ or $Cx \le d$) when $true$.
\vskip 6pt
\begin{equation}
    \label{eq:ex}
    \begin{bmatrix}
    Y_1 \\ Ax \leq b
    \end{bmatrix}
    \lor
    \begin{bmatrix}
        Y_2 \\ Cx \leq d
    \end{bmatrix}
\end{equation}
\begin{equation}
    \label{eq:x}
    0 \leq x \leq U
\end{equation}
\begin{equation}
    \label{eq:simple_xor}
    Y_1 \ \underline{\vee} \ Y_2
\end{equation}
\begin{equation}
    \label{eq:y}
    Y_1, Y_2 \in \{true, false\}
\end{equation}
\vskip 6pt

@@

 \subsubsection{Big-M Reformulation}
 The Big-M reformulation for this problem is given by \eqref{eq:ex_bigm1} - \eqref{eq:ex_bigm5}, where $M$ is a sufficiently large scalar that makes the particular constraint redundant when its indicator variable is not selected (i.e., $y_i = 0$). Note that the Boolean variables, $Y_i$, are replaced by binary variables, $y_i$. When the integrality constraint in Eq. \eqref{eq:ex_bigm4} is relaxed to $0 \leq x_1, x_2 \leq 1$, the resulting feasible region can be visualized by projecting the relaxed model onto the $x_1, x_2$ plane. This results in the region encapsulated by the dashed line in Figure \ref{fig:bigm}. It should be noted that the relaxed feasible region is not as tight as possible around the original feasible solution space. The choice of the large $M$ value determines the tightness of this relaxation, and the minimal value of $M$ for the optimal relaxation can be found through interval arithmetic when the model is linear. For nonlinear models, the tightest $M$ can be obtained by solving the maximization problem $\{\max h_{ik}(x): x \in X\}$. An alternate method for tight Big-M relaxations is given in \cite{TRESPALACIOS201598}.

\begin{equation}
    \label{eq:ex_bigm1}
    Ax \leq b + M \cdot (1 - y_1)
\end{equation}
\begin{equation}
    \label{eq:ex_bigm2}
    Cx \leq d + M \cdot (1 - y_2)
\end{equation}
\begin{equation}
    \label{eq:ex_bigm3}
    y_1 + y_2 = 1
\end{equation}
\begin{equation}
    \label{eq:ex_bigm4}
    0 \leq x \leq U
\end{equation}
\begin{equation}
    \label{eq:ex_bigm5}
    y_1, y_2 \in \{0,1\}
\end{equation}

@@

\subsubsection{Hull Reformulation}
The Hull reformulation is given by \eqref{eq:ex_hull0} - \eqref{eq:ex_hull6}, which requires lifting the model to a higher-dimensional space. When projected to the original space, the continuous relaxation of the model is tighter than its Big-M equivalent \cite{grossmann_trespalacios_2013}. The reformulation relaxation can be visualized by the region encapsulated by the dashed line in Figure \ref{fig:chr}. Note that this reformulation provides a tighter relaxation than the Big-M reformulation in Figure \ref{fig:bigm}. Also note that describing the geometry of this relaxation is more complex than the Big-M relaxation, which is made possible by the increased number of constraints and variables in the model.

\begin{equation}
    \label{eq:ex_hull0}
    Ax_1 \leq by_1
\end{equation}
\begin{equation}
    \label{eq:ex_hull1}
    Cx_2 \leq dy_2
\end{equation}
\begin{equation}
    \label{eq:ex_hull2}
    x = x_1 + x_2
\end{equation}
\begin{equation}
    \label{eq:ex_hull4}
    y_1 + y_2 = 1
\end{equation}
\begin{equation}
    \label{eq:ex_hull5}
    0 \leq x \leq U
\end{equation}
\begin{equation}
    \label{eq:ex_hull3}
    0 \leq x_i \leq U y_i \quad \forall i \in \{1,2\}
\end{equation}
\begin{equation}
    \label{eq:ex_hull6}
    y_1, y_2 \in \{0,1\}
\end{equation}

@@

Once the logic propositions are converted to CNF, each clause can be converted into an algebraic constraint with the following equivalence, where the set $I$ represents the subset boolean variables present in the clause, and the set $J$ represents the subset of boolean variables present in the clause in negated form,

\begin{align*}
    \bigvee_{i \in I} Y_i \bigvee_{j \in J} (\neg Y_j) & \ \ \text{becomes} \ \ \sum_{i \in I} y_i + \sum_{j \in J} (1-y_j) \geq 1 \\
\end{align*}

@@

\begin{lstlisting}[language = Julia]
using DisjunctiveProgramming, HiGHS

# create model
m = GDPModel(HiGHS.Optimizer)
# add variables to model
@variable(m, 0 <= F[i = 1:7] <= 10)
@variable(m, 0 <= CS <= CSmax)
@variable(m, CRmin <= CR <= CRmax)
# add logical variables to model
@variable(m, YR[1:2], LogicalVariable)
@variable(m, YS[1:2], LogicalVariable)

# add global constraints to model
@constraints(m,
    begin
        F[1] == F[2] + F[3]
        F[7] == F[5] + F[6]
    end
)
\end{lstlisting}
    \item Define the inner (nested) disjunction for the separation technologies in the superstructure using the \verb|@disjunction| macro.   
\begin{lstlisting}[language = Julia]
# define constraints in left YS disjunct
@constraint(m, F[5] == β[:S1]*F[4], DisjunctConstraint(YS[1]))
@constraint(m, CS == γ[:S1], DisjunctConstraint(YS[1]))
# define constraints in right YS disjunct
@constraint(m, F[5] == β[:S2]*F[4], DisjunctConstraint(YS[2]))
@constraint(m, CS == γ[:S2], DisjunctConstraint(YS[2]))
# define disjunction (specify parent disjunct)
@disjunction(m, YS, DisjunctConstraint(YR[2]))
\end{lstlisting}
    \item Define the outer disjunctions.
\begin{lstlisting}[language = Julia]
# define constraints in left YR disjunct
@constraint(m, F[6] == β[:R1]*F[2], DisjunctConstraint(YR[1]))
@constraint(m, [i = 3:5], F[i] == 0, DisjunctConstraint(YR[1]))
@constraint(m, CR == γ[:R1], DisjunctConstraint(YR[1]))
@constraint(m, CS == 0, DisjunctConstraint(YR[1]))
# define constraints in right YR disjunct
@constraint(m, F[6] == β[:R2]*F[3], DisjunctConstraint(YR[2]))
@constraint(m, CR == γ[:R2], DisjunctConstraint(YR[2]))  
# define disjunction
@disjunction(m, YR)    
\end{lstlisting}
    \item Add the selection logical constraints using the \verb|choose!| function. The first constraint enforces that only one reactor is selected (i.e., $Y_{R_1} \ \underline{\vee} \ Y_{R_2}$). The second constraint enforces that the separation system be defined only if the second reactor ($R_2$) is selected. This constraint is equivalent to the proposition $Y_{R_2} \Leftrightarrow Y_{S_1} \ \underline{\vee} \ Y_{S_2}$.
\begin{lstlisting}[language = Julia]
@constraint(m, YR in Exactly(1))
@constraint(m, YS in Exactly(YR[2]))
\end{lstlisting}
    \item Add the objective function and optimize. 
\begin{lstlisting}[language = Julia]
@objective(m, Max, F[7] - CS - CR)
optimize!(m)
\end{lstlisting}
\end{enumerate}
